Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2024-01-04 analysis/paper/discussion.qmd
fix: convert cross-refs to quarto and add figures and tables

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossil2010], which could in part be due to the 
lack of oral amylase activity in our model. 
<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardPlantMicroremains2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
who found that plants producing starches 10--20 $\mu$m in
size were over-represented; however, the representation of granules larger than
20 $\mu$m in their study is unclear.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranules2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossil2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
-->
The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculus2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodPathways2017; @radiniMedievalWomen2019].  
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are 
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to 
tooth surfaces (including plaque and dental calculus) and mucous membranes 
[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
Bacteria also have the ability to adhere to starch granules 
[@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by $\alpha$-amylase [@kashketFoodParticles1996].
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation, 
while dehydrated starches will have a reduced susceptibility
[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
@bjorckStarchProcessing1984; @henryCookingStarch2009].
Cracks on the surface of the dental calculus, as well as unmineralised islands 
and channels may also be able to contain starch granules 
[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
Starch granules that are trapped in these pockets are (at least to some extent) 
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus. 
The size bias against large granules (>20 $\mu$m) from both wheat and potato 
(Table \@ref(tab:sample-prop-tbl)) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
observed clusters of starches within dental calculus, rather than an even 
distribution across the surface of the dental calculus.
Granules trapped 
in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
After the death of an individual, starches within dental calculus are susceptible 
to further degradation by post-depositional processes, depending on burial environment 
(pH, temperature, moisture content, microorganisms)
[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
Future study should explore how burial affects the recovery of starch from the
biofilm model. <!-- I may get to it eventually, but it would be great if someone else could look at this! -->

>>>>

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus (@tbl-solution-count and @tbl-sample-count).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossil2010], which could in part be due to the 
lack of oral amylase activity in our model. 
<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardPlantMicroremains2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
who found that plants producing starches 10--20 $\mu$m in
size were over-represented; however, the representation of granules larger than
20 $\mu$m in their study is unclear.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(@tbl-sample-prop, and @fig-ratio-plots).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranules2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
retained starch granules ([@fig-cor-plot]A), meaning larger calculus deposits
contain a higher quantity of granules; a result that contradicts findings from
archaeological contexts
[@wesolowskiEvaluatingMicrofossil2010; @dudgeonDietGeography2014].
When the concentration of starch granules per mg calculus is considered, the
correlation is weaker, but still present ([@fig-cor-plot]B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
-->
The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculus2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodPathways2017; @radiniMedievalWomen2019].  
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are 
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to 
tooth surfaces (including plaque and dental calculus) and mucous membranes 
[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
Bacteria also have the ability to adhere to starch granules 
[@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by $\alpha$-amylase [@kashketFoodParticles1996].
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation, 
while dehydrated starches will have a reduced susceptibility
[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
@bjorckStarchProcessing1984; @henryCookingStarch2009].
Cracks on the surface of the dental calculus, as well as unmineralised islands 
and channels may also be able to contain starch granules 
[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
Starch granules that are trapped in these pockets are (at least to some extent) 
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus. 
The size bias against large granules (>20 $\mu$m) from both wheat and potato 
(@tbl-sample-prop) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
observed clusters of starches within dental calculus, rather than an even 
distribution across the surface of the dental calculus.
Granules trapped 
in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
After the death of an individual, starches within dental calculus are susceptible 
to further degradation by post-depositional processes, depending on burial environment 
(pH, temperature, moisture content, microorganisms)
[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
Future study should explore how burial affects the recovery of starch from the
biofilm model. <!-- I may get to it eventually, but it would be great if someone else could look at this! -->

### 2024-01-04 analysis/paper/matmet.qmd
rename sections to .qmd
fix: convert cross-refs to quarto and add figures and tables

## Biofilm formation

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDeposition1987; @sissonsMultistationPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).

@@

## Amylase activity detection

@@

## Treatment solutions

@@

## Extraction method

@@

## Statistical analysis

>>>>

### Biofilm formation

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDeposition1987; @sissonsMultistationPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see @fig-protocol for an overview of the protocol).

```{r}
#| label: fig-protocol
#| fig-cap: "Overview of experiment protocol including the plate setup."
knitr::include_graphics(here("figures/protocol_overview.png"))
```

@@

### Amylase activity detection

@@

### Treatment solutions

@@

### Extraction method

@@

### Statistical analysis

### 2024-01-04 analysis/paper/results.qmd
fix: convert cross-refs to quarto and add figures and tables

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
 analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

@@

A one-way ANOVA suggests that the type of starch used during the biofilm 
growth period had a minimal effect on the growth of the biofilm 
(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
= `r treat_f.stat`, p = `r treat_p`.
A summary of sample weights is available in Table \@ref(tab:anova-tbl).

@@

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
and in the biofilm samples 
(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

@@

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \@ref(tab:sample-prop-tbl)). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e., potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

@@

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figure \@ref(fig:ratio-plots)).

<!-- Proportion and size ratios are the most important points to be made -->

>>>>

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis (@fig-microscope), resulting in a total of 48 biofilm
samples (two plates of 24), 45 of which were used for analysis (three samples
were set aside for later analysis). Most control samples contained no starch
granules, while some contained negligible quantities (see Supplementary Material).

::: {#fig-microscope layout-ncol=2}

```{r}
knitr::include_graphics(here("figures/starches_w_bar.jpg"))#,
                          #here("figures/st2C3.2-mix.jpg"),
                          #here("figures/st1B4-wheat.jpg"),
                          #here("figures/2D2-potato.jpg")))
```

```{r}
knitr::include_graphics(here("figures/st2C3.2-mix.jpg"))
```

```{r}
knitr::include_graphics(here("figures/st1B4-wheat.jpg"))
```

```{r}
knitr::include_graphics(here("figures/2D2-potato.jpg"))
```

Microscope images of biofilm samples that were exposed to the starch solutions.
Starch granules can be seen within bacterial communities and isolated. Scale bar = 20 μm.

:::

@@

A one-way ANOVA suggests that the type of starch used during the biofilm 
growth period had a minimal effect on the growth of the biofilm 
(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
= `r treat_f.stat`, p = `r treat_p`.
A summary of sample weights is available in @tbl-anova.

```{r}
#| label: tbl-anova
#| tbl-cap: "Summary statistics for biofilm dry-weights (in mg) by treatment."
knitr::kable(anova_tbl, 
             col.names = c("Treatment", "Mean", "SD", "Min", "Max"))
```

@@

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
and in the biofilm samples 
(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
(@tbl-solution-count and @tbl-sample-count).

```{r}
#| label: tbl-solution-count
#| tbl-cap: "Mean starch counts from solutions, including the proportional makeup of the different sizes of granules."
cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
options(knitr.kable.NA = "")
knitr::kable(sol_tbl, col.names = cnames)
```

```{r sample-count-tbl}
#| label: tbl-sample-count
#| tbl-cap: "Mean starch counts extracted from samples with standard deviation (SD), including the proportion of granule sizes of the total count."
cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
            "Large (%)", "SD", "Total (%)", "SD")
knitr::kable(samp_counts_tbl, col.names = cnames)
```

@@

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples (@tbl-sample-prop). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e., potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

```{r}
#| label: tbl-sample-prop
#| tbl-cap: "The mean percentage of starches from the solutions that were incorporated into the samples."
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc_tbl, col.names = cnames)
```

@@

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(@fig-ratio-plots).

<!-- Proportion and size ratios are the most important points to be made -->
```{r}
#| label: fig-ratio-plots
#| fig-cap: "Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```
